---
title: "Weather Events"
author: "Mahesh Nepal"
date: "March 23, 2016"
---

Synopsis
---

The purpose of this analysis is to answer the following questions:

1.Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2.Across the United States, which types of events have the greatest economic consequences?


DataProcessing
---

First, the data is loaded into the workspace. Here, we load only the required columns for our analysis.
```{r}
storm_data <- read.csv(bzfile("repdata-data-StormData.csv.bz2"), header = T)
storm_data_lean <- storm_data[, c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG")]

storm_data_lean$EVTYPE <- gsub("^HEAT$", "EXCESSIVE HEAT", storm_data_lean$EVTYPE)
storm_data_lean$EVTYPE <- gsub("^TSTM WIND$", "THUNDERSTORM WIND", storm_data_lean$EVTYPE)
storm_data_lean$EVTYPE <- gsub("^THUNDERSTORM WIND$", "THUNDERSTORM WINDS", storm_data_lean$EVTYPE)
```

The following snippet shows the aggregate of the data of events that cause the **Fatalities**, **Injuries** and **Property Damage** respectively.

**Aggregate Fatalities**
```{r}
## aggregate fatalities
agg_fatalities <- aggregate(storm_data_lean$FATALITIES, by = list(storm_data_lean$EVTYPE), FUN = sum, na.rm = T)
colnames(agg_fatalities) <- c("event_type", "total_fatality")
sorted_fatalities <- agg_fatalities[order(-agg_fatalities$total_fatality),]
top_fatalities <- sorted_fatalities[1:10, ]
top_fatalities$event_type <- factor(top_fatalities$event_type, levels = top_fatalities$event_type, ordered = T)
```

**Aggregate Injuries**
```{r}
## aggregate injuries
agg_injuries <- aggregate(storm_data_lean$INJURIES, by = list(storm_data_lean$EVTYPE), FUN=sum, na.rm = T)
colnames(agg_injuries) = c("event_type", "total_injury")
sorted_injuries <- agg_injuries[order(-agg_injuries$total_injury),] 
top_injuries <- sorted_injuries[1:10,]
top_injuries$event_type <- factor(top_injuries$event_type, levels = top_injuries$event_type, ordered = T)
```

**Aggregate Property Damage**
```{r}
## aggregate property damage
agg_property_damage <- aggregate(storm_data_lean$PROPDMG, by = list(storm_data_lean$EVTYPE), FUN = sum, na.rm = T)
colnames(agg_property_damage) = c("event_type", "total_property_damage")
sorted_property_damage <- agg_property_damage[order(-agg_property_damage$total_property_damage),] 
top_property_damage <- sorted_property_damage[1:10,]
top_property_damage$event_type <- factor(top_property_damage$event_type, levels = top_property_damage$event_type, ordered = T)
```


Results
---
The results of the analysis are being illustrated by the following plot

**Fatalities**
```{r}
library(ggplot2)
ggplot(data=top_fatalities, aes(x=event_type, y=total_fatality)) + 
    geom_bar(stat="identity") + xlab("Event type") + ylab("Total fatalities") + 
    ggtitle("Fatalities By Event Type") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Injuries**
```{r}
ggplot(data=top_injuries, aes(x=event_type, y=total_injury)) + 
    geom_bar(stat="identity") + xlab("Event type") + ylab("Total injuries") + 
    ggtitle("Injuries By Event Type") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Property Damage**
```{r}
ggplot(data=top_property_damage, aes(x=event_type, y=total_property_damage)) + 
    geom_bar(stat="identity") + xlab("Event type") + 
    ylab("Total property damage") +  ggtitle("Property Damage By Event Type") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```